---
title: "DRAFT HW 9: BSTA 511-611 F23"
author: "Your name here - update this!!!!"
date: "Due 12/2/23"
format: 
  html:
    link-external-newwindow: true
    toc: true
    toc_float:
      collapsed: false
    number-sections: true
    number-depth: 3
    theme: lumen
    self-contained: true
    html-math-method: mathjax
---

```{r}
#| label: "setup"
#| include: false
knitr::opts_chunk$set(echo = TRUE)
```


**Due 12/2/23**

Download the .qmd file for this assignment from [https://github.com/niederhausen/BSTA_511_F23/blob/main/homework/HW_8_F23_bsta511.qmd](https://github.com/niederhausen/BSTA_511_F23/blob/main/homework/HW_8_F23_bsta511.qmd)


## Directions {.unnumbered}

* Please upload your homework to Sakai. __Upload both your .qmd code file and the rendered .html file__.
* For each question, make sure to include all code and resulting output in the html file to support your answers. 


### R & LaTeX code {.unnumbered}

* See the .qmd files with the code from class notes for LaTeX and R code.
* The LaTeX code will make it easier to show your work in computations.

:::{.callout-tip}
It is a good idea to try rendering your document from time to time as you go along!  Note that rendering automatically saves your Qmd file and rendering frequently helps you catch your errors more quickly.
:::


# Book exercises {.unnumbered}


## 6.12 Trends in the residuals {.unnumbered}

## 6.20 Guppies, Part IV {.unnumbered}

## 6.26 (a, e) Helmets and lunches {.unnumbered}

__Skip parts (b)-(d). To complete (e), use that the slope from part (b) is −0.537, and the intercept is 55.34.__

_Note: if you have time, it would be good practice to calculate the regression line as well. This was covered on the previous assignment._ 



## 6.28 Guppies, Part V {.unnumbered}

## 6.32 (a, b) Guppies, Part VI  {.unnumbered}

__Skip part (c).__





# R exercises 

## Load all the packages you need below here. 

```{r}

```


## R1: Coming soon! 



