%\documentclass[12pt,letterpaper]{article}
\documentclass[12pt]{amsart}

\usepackage[left=1in, right=1in, top=.5in, bottom=.5in]{geometry}
\usepackage{latexsym,amssymb,amsmath,amsthm,amsopn,verbatim, mathpazo, graphicx, lmodern, hyperref, color}
%\usepackage{latexsym,amssymb,amsmath,amsthm,amsopn,verbatim, mathpazo, graphicx}
\newcommand{\vs}{\vskip.5cm}
\newcommand{\hs}{\hskip1cm}
\newcommand{\ds}{\displaystyle}
\setlength{\parindent}{0pt}


\usepackage[T1]{fontenc}
\usepackage{libertine}
\renewcommand*\familydefault{\sfdefault}  %% Only if the base font of the document is to be sans serif


\newtheorem{theorem}{Theorem}[section]
\newtheorem{corollary}{Corollary}[theorem]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{example}[theorem]{Example}

\newcommand\indep{\protect\mathpalette{\protect\independenT}{\perp}}
\def\independenT#1#2{\mathrel{\rlap{$#1#2$}\mkern2mu{#1#2}}}

\begin{document}


%----------------------------------------------------------------------------
\setcounter{section}{3}
%\setcounter{subsection}{1}
%\setcounter{theorem}{1}
Day 7 BSTA 511/611
{\huge  
\section*{Chapter 3: Distributions of Random Variables (Part 1)}
}

%----------------------------------------------------------------------------


%----------------------------------------------------------------------------
{\large 
%----------------------------------------------------------------------------


%----------------------------------------------------------------------------
%\begin{example}  \textbf{text.} \newline
%
%\begin{enumerate}
%\item 
%
%\vspace{2cm}
%
%\item 
%
%\vspace{2cm}
%
%\item 
%
%\vspace{2cm}
%\end{enumerate}
%\end{example} 

%\newpage
%----------------------------------------------------------------------------

%Section 3.1
\subsection{Random variables}

\begin{definition} A \textbf{random variable (r.v.)} assigns numerical values to the outcome of a random phenomenon.
\end{definition}

Notation: \newline 
A random variable is usually denoted with a capital letter such as $X$, $Y$, or $Z$.

\vspace{1cm}
%----------------------------------------------------------------------------
\begin{example}  \textbf{Data points} \newline
Suppose you have a dataset of size 3 ($n=3$) with $x_1 = 5, x_2 = 3,$ and $x_3 = 6$. 
\begin{itemize}
\item Each data point is the outcome of a random phenomenon

\item Each data point is a numerical value
\item The data points are examples of values of a sequence of random variables $X_1, X_2,$ and $X_3$
\item For datasets, we almost always assume the data points came from random variables that are independent and have the same \textbf{distribution}. 
\item To calculate the likelihood of data, we need to know the distribution of the random variable that models the data.

\item First, let's remind ourselves how to calculate them mean and variance of a dataset:
	\begin{itemize}
	\item What is the mean of the data points?

%delete
\color{blue}
$$\bar{x} = 14/3 = 4.66667$$
\color{black}
\vspace{1.5cm}
	\item What are the variance and standard deviation of the data points?

%delete
\color{blue}
$$
s^2 = \frac{\sum_{i=1}^n(x_i - \bar{x})^2}{n-1}
$$
$$s^2 = 2.3333$$
$$s = 1.527525 \sim 1.53$$
\color{black}

	\end{itemize}


\end{itemize}
\end{example} 

\newpage
%----------------------------------------------------------------------------

%3.1.1
%\subsubsection{Distributions of random variables}


%----------------------------------------------------------------------------
\begin{example}  \textbf{Rolling a die} \newline
Suppose you roll a fair die. Let the random variable (r.v.) $X$ be the outcome of the roll, i.e. the value of the face showing on the die. 

\begin{enumerate}
\item What is the probability distribution of the r.v. $X$?
\vspace{3cm}

\item  What is the expected outcome of the r.v. $X$?

%delete
\color{blue}
\begin{itemize}
\item $$7/2 = 3.5$$
\item Not a possible outcome! 
\item Do not round.
\end{itemize}

\color{black}


\vspace{4cm}

\item Now suppose the 6-sided die is not fair. How would we calculate the expected outcome?

%%undelete
%\vspace{1cm}
%\begin{tabular}{| c | c |}
%  \hline                       
%  x & $\mathbb{P}(X=x)$  \\
%   \hline     
%  1 & 0.10 \\
%  2 & 0.20  \\
%  3 & 0.05  \\
%  4 & 0.05  \\
%  5 & 0.25  \\
%  6 & 0.35  \\
%  \hline  
%\end{tabular}


%delete
\color{blue}
\vspace{1cm}
\begin{tabular}{| c | c | c |}
  \hline                       
  $x$ & $\mathbb{P}(X=x)$ & $x\mathbb{P}(X=x)$ \\
   \hline     
  1 & 0.10 & 0.1\\
  2 & 0.20  & 0.4\\
  3 & 0.05  & 0.15\\
  4 & 0.05  & 0.2\\
  5 & 0.25  & 1.25\\
  6 & 0.35  & 2.10\\
   \hline   
   sum & 1 & $\mu = 4.2$ \\
  \hline  
\end{tabular}
\color{black}
%delete

\vspace{2cm}
\end{enumerate}
\end{example} 

\newpage
%----------------------------------------------------------------------------


(From $\S$ 2.1.5)
\begin{definition} A \textbf{probability distribution} consists of all disjoint outcomes and their associated probabilities.
\end{definition}

\textbf{Rules for a probability distribution} \newline
A probability distribution is a list of all possible outcomes and their associated probabilities that satisfies three rules: 
\begin{enumerate}
\setlength{\itemsep}{0mm}
\item The outcomes listed must be disjoint.
\item Each probability must be between 0 and 1.
\item The probabilities must total to 1. \vspace{1mm}
\end{enumerate}

\vspace{0.5cm}

Probability distributions are usually either \textbf{discrete} or \textbf{continuous}, depending on whether the random variable is discrete or continuous.

\vspace{1cm}

(Back to $\S$ 3.1.1)
\begin{definition} A \textbf{discrete} r.v. $X$ takes on a finite number of values or countably infinite number of possible values.
\end{definition}

\begin{definition} A \textbf{continuous} r.v. $X$ can take on any real value in an interval of values or unions of intervals.\end{definition}


\vspace{2cm}
%delete
\color{blue}
\textbf{Class notes:}

Sketch histogram of discrete values 1, 2, ..., 6 with continuous approximation

X could be values from a die or a waiting time in minutes

\color{black}

\newpage
%----------------------------------------------------------------------------

$\S$ \textbf{3.1.2 Expectation}

\vspace{.5cm}

\begin{itemize}
\item We call the mean of a random variable its \textbf{expected value}
\item The expected value is calculated as a weighted average
\end{itemize}

\vspace{.5cm}

\begin{definition}{\textbf{Expected value} of a discrete random variable} \newline
If $X$ takes on outcomes $x_1$, ..., $x_k$ with probabilities $P(X=x_1)$, ..., $P(X=x_k)$, the expected value of $X$ is the sum of each outcome multiplied by its corresponding probability:

%delete
\color{blue}
\begin{align}
\mu = E(X) 	&= x_1 P(X=x_1) + \cdots + x_k P(X=x_k) \notag \\
&= \sum_{i=1}^{k}x_iP(X=x_i). \notag 
\end{align}
\color{black}
\end{definition}

\vspace{1.5cm}

$\S$ \textbf{3.1.3 Variability of random variables}

\vspace{.5cm}

\begin{itemize}
\item Just like with data, the variability of a r.v. is described with its variance or standard deviation.
\item Squared deviations from the mean are weighted by their respective probabilities
\end{itemize}

\vspace{.5cm}

\begin{definition}{\textbf{Variance} of a discrete random variable} \newline
If $X$ takes on outcomes $x_1$, ..., $x_k$ with probabilities $P(X=x_1)$, \dots, $P(X=x_k)$ and expected value $\mu=E(X)$, then the variance of $X$, denoted by $\text{Var}(X)$ or $\sigma^2$, is

%delete
\color{blue}
\begin{align}
Var(X) &= (x_1-\mu)^2 P(X=x_1) + \cdots \notag + (x_k-\mu)^2 P(X=x_k) \notag \\
&= \sum_{i=1}^{k} (x_i - \mu)^2 P(X=x_i). \notag 
\end{align}
\end{definition}
\color{black}
\vspace{2cm}

\begin{definition}{\textbf{Standard deviation} of a discrete random variable} \newline
The standard deviation of $X$, labeled $SD(X)$ or $\sigma$, is 

%delete
\color{blue}
$$ SD(X) = \sigma = \sqrt{\sigma^2}$$
\color{black}

\end{definition}




%----------------------------------------------------------------------------
\begin{example}  \textbf{Rolling a fair die: variance} \newline
Suppose you roll a fair 6-sided die. Let the random variable (r.v.) $X$ be the outcome of the roll, i.e. the value of the face showing on the die.  \newline
Find the variance and standard deviation of $X$.


%%undelete
%\vspace{1cm}
%\begin{tabular}{| c | c |}
%  \hline                       
%  x & $\mathbb{P}(X=x)$  \\
%   \hline     
%  1 & 0.10 \\
%  2 & 0.20  \\
%  3 & 0.05  \\
%  4 & 0.05  \\
%  5 & 0.25  \\
%  6 & 0.35  \\
%  \hline  
%\end{tabular}

%delete
\vspace{1cm}

\color{blue}

$$\mu = 3.5$$

\vspace{1cm}
\begin{tabular}{| c | c | c | c | c | c |}
  \hline                       
  $x$ & $\mathbb{P}(X=x)$ &  $x - \mu$ & $(x - \mu)^2$ & $\mathbb{P}(X=x)(x - \mu)^2$  R\\
   \hline     
  1 & 1/6  & -2.5 &  6.25 & 1.04166667\\
  2 & 1/6    & -1.5 & 2.25 &  0.3750\\
  3 & 1/6    & -0.5 &  0.25 & 0.04166667\\
  4 & 1/6    & 0.5 & 0.25 &  0.04166667\\
  5 & 1/6   & 1.5   &  2.25 & 0.3750\\
  6 & 1/6    &2.5 & 6.25 & 1.04166667\\
   \hline   
   sum & 1 &    &  & $\sigma^2 = 2.916667$\\
  \hline  
\end{tabular}
\vspace{1cm}
%delete
$$\sigma =1.707825 $$


\vspace{1cm}
R commands


\begin{tabular}{| c | c | c | c | c | c |}
  \hline                       
  $x$ & $\mathbb{P}(X=x)$ &  $x - \mu$ & $(x - \mu)^2$ & $\mathbb{P}(X=x)(x - \mu)^2$  R\\
   \hline     
 $x <- 1:6$  & $px <- 1/6$  & $x - mu$ &  $(x - mu)^2$ & $px*(x - mu)^2$\\
   \hline   
   sum & 1 &    &  & $sum(px*(x - mu)^2)$\\
  \hline  
\end{tabular}
\vspace{2cm}


\end{example} 



 \newpage
%----------------------------------------------------------------------------

%----------------------------------------------------------------------------
\begin{example}  \textbf{Vaccinated people testing positive for Covid-19} \newline
About 25\% of people that test positive for Covid-19 are vaccinated for Covid-19.\newline
Define the r.v. $X$ to be 1 if someone that tests positive is vaccinated and 0 if they are not vaccinated.


\begin{enumerate}
\item Make a table for the probability distribution for the r.v. $X$

\vspace{3cm}

\item What is the expected value of $X$?

\vspace{2cm}
%delete
\color{blue}
$$\mathbb{E}[X] = p = 0.25$$
\color{black}

\item What is the variance of $X$?

\vspace{2cm}
%delete
\color{blue}
$$Var[X] = pq = 0.25\cdot0.75 = 0.1875$$
$$SD[X] = 0.4330127$$

\color{black}
\end{enumerate}

%\vspace{1cm}

%\vspace{1cm}
\end{example} 





\newpage
%----------------------------------------------------------------------------


$\S$ \textbf{3.1.4 Linear combinations of random variables}

%\vspace{1cm}
%delete
\color{blue}
Seems abstract, but actually fundamental to basic statistical theory - using it all the time!
\color{black}

%----------------------------------------------------------------------------
%\vspace{.5cm}

\begin{definition}{\textbf{Linear combinations} of random variables.} \newline
If $X$ and $Y$ are random variables and $a$ and $b$ are constants, then 
%delete
\color{blue}
$$aX+bY$$
\color{black}
\vspace{1cm}

is a linear combination of the random variables.
\end{definition}


%\vspace{1.5cm}

%----------------------------------------------------------------------------
\begin{theorem}{\textbf{Expected value of a linear combination} of random variables.} \newline
If $X$ and $Y$ are random variables and $a$ and $b$ are constants, then 
%delete
\color{blue}
$$\mathbb{E}(aX + bY) = a\mathbb{E}(X) + b\mathbb{E}(Y) = a\mu_X + b\mu_Y$$
\color{black}
\vspace{1cm}


\end{theorem}



%%----------------------------------------------------------------------------
%\begin{example}  \textbf{Rolling 3 dice} \newline
%Suppose you roll 3 fair 6-sided dice.  \newline
%Let the random variables $X_1, X_2, X_3$ be the values shown on the 3 dice.  \newline
%What is the expected sum of the 3 dice?
%
%\vspace{1cm}
%
%%delete
%\color{blue}
%
%\color{black}
%
%\vspace{1cm}
%\end{example} 



%----------------------------------------------------------------------------
\begin{example}  \textbf{Expected money for rolling 3 dice} \newline
Let the random variables $X_1, X_2, X_3$ be the values shown on 3 fair 6-sided dice rolls. \newline 
Suppose you are given in dollars the amount of the first roll, plus twice the value of the second roll, plus 4 times the value of the third roll. \newline 
How much money do you expect to get?  \newline 
%What is the standard deviation of the amount?

%\vspace{1cm}

%delete
\color{blue}
$$\mathbb{E}(X_1 + 2X_2 + 4X_3) = \mathbb{E}(X_1) + 2\mathbb{E}(X_2) + 4\mathbb{E}(X_3) = 7\cdot 3.5 = 24.5$$
\color{black}

\end{example} 

\vspace{2cm}

%delete
\color{blue}
Expected value of the sample mean
\color{black}


\newpage
%----------------------------------------------------------------------------
\begin{theorem}{\textbf{Variance of a linear combination} of random variables.} \newline
If $X$ and $Y$ are INDEPENDENT random variables and $a$ and $b$ are constants, then 
%delete
\color{blue}
$$\text{Var}(aX + bY) = a^2 \text{Var}(X) + b^2\text{Var}(Y)$$
\color{black}
\vspace{.5cm}


\end{theorem}

%----------------------------------------------------------------------------
\begin{example}  \textbf{Variance of money for rolling 3 dice} \newline
Let the random variables $X_1, X_2, X_3$ be the values shown on 3 fair 6-sided dice rolls. \newline 
Suppose you are given in dollars the amount of the first roll, plus twice the value of the second roll, plus 4 times the value of the third roll. \newline 
What are the variance and standard deviation of the amount you get from the 3 rolls?  \newline 


%\vspace{1cm}

%delete
\color{blue}
$$\text{Var}(X_1 + 2X_2 + 4X_3) = \text{Var}(X_1) + 4\text{Var}(X_2) + 16\text{Var}(X_3) = 21\cdot 17.5 = 367.5$$
\vspace{1cm}
$$\text{SD}(X_1 + 2X_2 + 4X_3) = \sqrt{367.5} =19.17029 $$
\color{black}
%\vspace{1cm}
\end{example} 



%delete
\vspace{1cm}
\color{blue}
Variance of the sample mean
\color{black}


 \newpage
%----------------------------------------------------------------------------

%----------------------------------------------------------------------------
\begin{example}\label{VaccBinom}  \textbf{Vaccinated people testing positive for Covid-19 (revisited)} \newline
About 25\% of people that test positive for Covid-19 are vaccinated for Covid-19.\newline
Define the r.v. $X_i$ to be 1 if someone that tests positive is vaccinated and 0 if they are not vaccinated. \newline
Suppose 3 people have tested positive for Covid-19 (independently of each other). \newline
Let $T$ denote the number of people that are vaccinated amongst the 3 that tested positive.


\begin{enumerate}
\item Using the r.v.'s $X_i$, write a mathematical equation for calculating $T$.

%\vspace{2cm}
%delete
\color{blue}
$$T = \sum_{i=1}^3 X_i$$
\color{black}

\item What is the expected value of $T$?

\vspace{1.5cm}
%delete
\color{blue}
$$\mathbb{E}[T] = 3\cdot p = 3\cdot0.25 = 0.75$$
\color{black}

\item What is the variance of $T$?

\vspace{2cm}
%delete
\color{blue}
$$Var[T] = npq = 3\cdot0.25\cdot0.75 = 3\cdot0.1875 = 0.5625$$
$$SD[T] = 0.5625$$

\color{black}


\item What is the probability distribution of $T$?

%delete
\color{blue}
\vspace{1cm}
\begin{tabular}{| c | c |}
  \hline                       
  x & $\mathbb{P}(X=x)$  \\
   \hline     
  0 & $0.75^3$ = 0.422\\
  1 & $ 3\cdot0.25\cdot0.75^2$ = 0.422\\
  2 & $ 3\cdot0.25^2\cdot0.75$  = 0.141\\
  3 &$0.25^3$ =  0.016\\
  \hline  
\end{tabular}
\color{black}

\end{enumerate}

%\vspace{1cm}

%\vspace{1cm}
\end{example} 




%----------------------------------------------------------------------------
}  % end large font
%----------------------------------------------------------------------------



\end{document}




